A free/open-source machine translation system for English to Kazakh

### 2015-03-05 engkaz-eamt-2015/engkaz-eamt-2015.tex
acknowl.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59219 72bbbca6-d526-0410-a7d9-f06f51895060
some fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59223 72bbbca6-d526-0410-a7d9-f06f51895060

% Authors:
% Aida, Mikel, Fran, Inari, ... ?
%Aida Sundetova
%Al-Farabi Kazakh National University, Al-Farabi av., 71, 050040 Almaty, Kazakhstan

@@

\begin{abstract}\todo{Decide authors, and put them in phantom boxes}
This paper presents the current state of development a shallow-transfer rule-based machine translation (MT) system from English to Kazakh. The main syntactic and morphological differences 
between the two languages are presented, and it is showed how the MT 
system was designed to tackle these challenges. We show an evaluation of system coverage and 
translation quality and outline and future work.
\end{abstract}

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based. 
Rule-based machine translation, in which developers 
write up dictionaries and grammatical rules that are applied by an engine, in some 
case, where statistics not work well, will produce better translation.
\todo{What does the last sentence mean? Clarify!}
This paper describes work in progress of development of machine translation system 
for English--Kazakh using the Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

\todo{I have removed some text and  the statistical PoS tagging stuff, as it is not used}

@@

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-лар}, \emph{-лер}, \emph{-дар}, \emph{-лер} depending on vowel harmony: кітап+тар,
мектеп+тер, etc. \todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; DONE  homogeneize notation in paper.}

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,135 stems \citep{sundetova13a}.\todo{later you say 13,135}

@@

\paragraph{Chunker:}\todo{This should have 4 numerals, but I don't know how to do this}
In the first round of structural transfer, rules segment sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into equivalent Kazakh chunks, leaving some adaptations to be performed in later stages of structural transfer (for instance, the 
morphological case of noun phrases). Table~\ref{table:nps} describes what kind of noun phrases could appear in sentences and how it could be detected.\todo{This table does not add much. the first and second column of the table are essentially the same: it would be more interesting to give the Kazakh sequence of PoS tags, perhaps with an example}

\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|}
  \hline
  \textbf{Structure} & \textbf{Description} & \textbf{Examples} \\
  \hline
  \texttt{n} & Noun & house \\
  \texttt{adj + n} & Adjective + noun & long road \\
  \texttt{det + n} & Determiner + noun & an apple, the chance \\
  \texttt{num + n} & Numeral + noun & five pieces \\
  \texttt{num + adj + n} & Numeral + adjective + noun & five big pieces \\
  \texttt{det + num + adj + n} & Determiner + numeral + adjective + noun & the five big pieces \\
  \hline
  \end{tabular}
  \caption{Example of the most common structures of noun phrases}
  \label{table:nps}
\end{table*}

@@

\todo{I removed text that pertained to kaz--eng}

@@

  English present continuous or past continuous (I am playing, I was playing), have to be detected and 
  mapped onto sets of two lexical units (Мен ойнап жатырмын, Мен ойнап отырдым) where the main verb is 
  found in the -п participle form (ойнап), and a suitable finite form  (жатырмын, отырдым) of an auxiliary 
  verb (жатыр, отыр) is used to express number and person agreement,\footnote{Actually, Kazakh language uses four auxiliary verbs: жатыр ('lie', 
      used when the activity takes a long time), отыр ('sit', used when the activity appears to be done in a sitting position), тұр ('stand', 
      when the takes a short time), and жүр (when the activity repeats regularly). Choosing the most adequate auxiliary verb is 

@@

      is жатыр ('lie') for the present continuous and отыр ('sit') for the past continuous.} 
  modal verbs are translated by adding adjectives, which means ``necessary'' (керек) or ``proper'' (жөн), \todo{All examples in italics. This part is hard to read. The choice of auxilliary is not justified.}

@@

MLF thanks the Kazakh state program for the attraction of foreign scholars and Prof. Ualsher Tukeyev for supporting his visit to the Kazakh National University, where part of this work was carried out. We also thank Prof. Tukeyev for his valuable input.\todo{Some references didn't come out right.}
\bibliographystyle{apalike}

>>>>

\author{Aida Sundetova\\
  Kazakh National University,  \\
  050040 Almaty, \\
  {\tt xxx@xxx}\\
  \And
  \textbf{Francis M. Tyers}\\
  HSL-fakultetet\\ 
  UiT Norgga \'{a}rktala\v{s} universitehta \\
  9017 Romsa, \\
  {\tt francis.tyers@uit.no}
  \And
  Mikel L. Forcada\\
  Dept. Lleng. i Sist. Inform. \\
  Universitat d'Alacant, \\ 
  Alacant \\
  {\tt mlf@dlsi.ua.es}  
}

@@

\begin{abstract}
This paper presents the current state of development a shallow-transfer rule-based machine 
translation (MT) system from English to Kazakh. The main syntactic and morphological differences 
between the two languages are presented, and it is showed how the MT 
system was designed to tackle these challenges. We show an evaluation of system coverage and 
translation quality and outline and future work.
\end{abstract}

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing 
commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} 
and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based.

This paper describes the development of a machine translation system 
for English to Kazakh using the Apertium free/open-source machine 
translation platform \citep{forcada11}. %(http://www.apertium.org).

@@

@@

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-лар}, \emph{-лер}, \emph{-дар}, \emph{-лер} depending on vowel harmony: кітап+тар,
мектеп+тер, etc. 
%\todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; DONE  homogeneize notation in paper.}

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,135 stems \citep{sundetova13a}.

@@

\paragraph{Chunker:}\todo{This should have 4 numerals, but I don't know how to do this}
In the first round of structural transfer, rules segment sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into equivalent Kazakh chunks, leaving some adaptations to be performed in later stages of structural transfer (for instance, the 
morphological case of noun phrases).

%Table~\ref{table:nps} describes what kind of noun phrases could appear in sentences and how it could be detected.\todo{This table does not add much. the first and second column of the table are essentially the same: it would be more interesting to give the Kazakh sequence of PoS tags, perhaps with an example}

% \begin{table*}
%   \centering
%   \begin{tabular}{|l|l|l|}
%   \hline
%   \textbf{Structure} & \textbf{Description} & \textbf{Examples} \\
%   \hline
%   \texttt{n} & Noun & house \\
%   \texttt{adj + n} & Adjective + noun & long road \\
%   \texttt{det + n} & Determiner + noun & an apple, the chance \\
%   \texttt{num + n} & Numeral + noun & five pieces \\
%   \texttt{num + adj + n} & Numeral + adjective + noun & five big pieces \\
%   \texttt{det + num + adj + n} & Determiner + numeral + adjective + noun & the five big pieces \\
%   \hline
%   \end{tabular}
%   \caption{Example of the most common structures of noun phrases}
%   \label{table:nps}
% \end{table*}

@@

@@

  English present continuous or past continuous (\emph{I am playing}, \emph{I was playing}), have to be detected and 
  mapped onto sets of two lexical units (\emph{Мен ойнап жатырмын}, \emph{Мен ойнап отырдым}) where the main verb is 
  found in the -п participle form (\emph{ойнап}), and a suitable finite form  (\emph{жатырмын}, \emph{отырдым}) of an auxiliary 
  verb (\emph{жатыр}, \emph{отыр}) is used to express number and person agreement,\footnote{Actually, Kazakh language uses four auxiliary verbs: \emph{жатыр} `lie', 
      used when the activity takes a long time), отыр (`sit', used when the activity appears to be done in a sitting position), \emph{тұр} `stand', 
      when the takes a short time), and \emph{жүр} (when the activity repeats regularly). Choosing the most adequate auxiliary verb is 

@@

      is \emph{жатыр} `lie' for the present continuous and отыр (`sit') for the past continuous.} 
  modal verbs are translated by adding adjectives, which means \emph{керек} `necessary' or \emph{жөн} `proper', \todo{This part is hard to read. The choice of auxilliary is not justified.}

@@

MLF thanks the Kazakh state program for the attraction of foreign scholars and Prof. Ualsher Tukeyev for supporting his visit to the Kazakh National University, where part of this work was carried out. We also thank Prof. Tukeyev for his valuable input. Special thanks to Inari Listenmaa for her assistance in mentoring the Google Summer of Code project.\todo{Some references didn't come out right.} 
\bibliographystyle{apalike}
